Runtime-Assisted Convergence in Replicated Data Types

### 2022-03-21 related.tex
Edits for shepherd review and camera-ready

The idea of replicated data types with three-way merge functions was
first proposed in~\cite{mrdt}, where authors demonstrate an approach
for deriving merge functions from first principles for various data
types. However, the resultant \emph{mergeable replicated data types}
(MRDTs) were never proved to be convergent. As we demonstrate for
the \C{Set} data type in Sec.~\ref{sec:motivation}, simple merge
functions with intuitive merge semantics can nonetheless lead to
divergent executions. \quark can thus be used with MRDTs to preempt
such executions and guarantee their convergence.

Concurrent Revisions~\cite{BBL+10} proposes versioned state for
shared-memory concurrent programs in order to guarantee determinism.
The fork-join concurrency they consider is however too restrictive for
the asynchronous distributed setting where replicas do not have a
single coherent view of the system state.  Furthermore the
hierarchical organization of the concurrent actors with a designated
\emph{server} is inconsistent with the fully-decentralized model of
execution adopted by several collaborative applications.

@@

>>>>

Mergeable Replicated Data Types (MRDTs) were introduced
in~\cite{mrdt}, where authors also demonstrate an approach for
deriving merge functions from first principles.  Despite resulting in
sensible merge semantics for several data types, their approach was
never shown to guarantee the convergence of the resultant MRDTs.
Indeed, as we demonstrate in Sec.~\ref{sec:motivation}, convergence
of unrestricted MRDT executions cannot be guaranteed due to the
presence of anomalies such as Fig.~\ref{fig:mrdt-exec-2} and
Fig.~\ref{fig:mrdt-exec-4}. \quark fixes this problem by extending
MRDTs with a runtime that limits their executions to those that are
guaranteed to converge.

% Concurrent Revisions~\cite{BBL+10} proposes versioned state for
% shared-memory concurrent programs in order to guarantee determinism.
% The fork-join concurrency they consider is however too restrictive for
% the asynchronous distributed setting where replicas do not have a
% single coherent view of the system state.  Furthermore the
% hierarchical organization of the concurrent actors with a designated
% \emph{server} is inconsistent with the fully-decentralized model of
% execution adopted by several collaborative applications.

@@

% Type systems
The meta-theory of \quark abstract machine
(Sec.~\ref{sec:abstract-sem} bears resemblance to the type safety
proofs carried out in Wright and Felleisen's style~\cite{WF92}.
Theorem~\ref{thm:progress} (Progress) is analogous to the Progress
lemma of type safety, and Theorem~\ref{thm:convergence} (Convergence)
is analogous to Preservation.  However, unlike the type-based
approaches, safety in \quark is enforced through run-time monitoring.
Static enforcement of convergence in MRDTs using, for e.g., Session
Types~\cite{HNMSession16}, is an interesting direction for future
work.

% Git
\quark's programming model is inspired by distributed version control
systems in general, and Git in particular. To the best of our
knowledge, operational semantics of Git has never been formalized.
This is not a serious concern considering that humans are heavily
involved in Git workflows, and the complexity of version histories
manifest by Git is bound by the limits of human cognition.
Nonetheless, it is possible to construct Git version histories that
result in non-convergent and counter-intuitive final
states~\cite{tycon,russell}. Such anomalies are more likely to occur
if Git version histories are manifest commensurately with distributed
executions of computer programs. \quark's run-time monitoring
pre-empts this possibility in our case.
